This thesis would not have been possible without my many communities, nor without the contributions of those who donated their data to further our understanding of genetics.

I would like to thank Simon Gravel for his guidance and mentorship and his balance of energetic curiosity with scientific scepticism, and the members of my supervisory committee, Nada Jabado and Ryan Hernandez. I would also like to thank the QLS program, particularly Celia Greenwood and Alex DeGuise for all of their hard work.

Thank you to my QLS cohort for grinding it out with me: Myriah Haggard, Matt D'Iorio, Jeffrey Hyacinthe, Yixiao Zheng, Sara Zapata-Marin, and Selin Jessa; to Noor Al-Sharif and Jake Vogel, for making Montreal home; to my Thomson House Trivia crew: Matt, Barbara, Hector, Gerardo, JT, Val, and Rozzy, for being champions; to my lab, especially my desk-mate Chief Ben-Eghan for cracking me up; to the flor\ae ~and faun\ae ~of Parc La Fontaine, for raising my spirits.

Thank you to my friends in my other homes. Those in Ottawa: Dylan, Leo, Julie, Ashley, Shane, Isaac, Amy, thanks for always hosting me when I was in town. To those in KW: Jake, Kayla, Warren, Kaitlin, Jon, Steph, Derek, Christina, Dave, Rebecca, Travis, Cynthia, Tyler, and Aine, thanks for entertaining my rants on Slack---may your broods grow belligerent and numerous.

Thank you to my loved ones: to Melissa, for your love and support and for exploring the forests and rivers with me; to my brother, Andrew, and my mother, Marina, for carrying me when I stumbled; and lastly to my father, Andr\'{e}s, who started this marathon with me so long ago---the finish line is in sight, and I can hear you cheering.

\pagebreak
\hspace{0pt}
\vfill

\begin{center}
%\begin{minipage}{.7\linewidth}
To my family; and, \\
To the memory of Andr\'{e}s Carlos D\'{i}az Bravo.
%\end{minipage}
\end{center}

\vfill
\hspace{0pt}
\pagebreak
